Component response rate variation underlies stability in large complex systems

### 2019-02-27 notebook/ms.Rmd
Edit the manuscript up to Reviewer 2 specific comment 3 -- these are next on the list, specifically the new Discussion paragraph

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($E[\gamma]$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) in which $\gamma$ varies are actually more likely to be stable.  <!--- Need to emphasise that this is for *real* systems at the thresholds of stability --->

@@

<!--- Also important to emphasise Gibbs result -- I'm doing this for finite systems, and I deliberate stressed the system complexity to the point of collapse. --->

@@

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} =\mathbf{\gamma} \mathbf{A}$ modulates the values of components and can be analysed using the techniques of May [@May1972; @May1973; @Ahmadian2015].

>>>>

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) in which $\gamma$ varies are actually more likely to be stable.  <!--- Need to emphasise that this is for *real* systems at the thresholds of stability --->

@@

Also important to emphasise Gibbs result -- I'm doing this for finite systems, and I deliberate stressed the system complexity to the point of collapse.

But Gibbs was more interested in first assuming a stable matrix and then showing that the vector of abundances would not change it as S \to infinity -- Instead of assuming the system was stable prior to adding component response rate, I allow both the system and component response rate to vary to see the effects on stability.

@@

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} = \mathbf{\gamma} \mathbf{A}$ modulates the values of components and can be analysed using the techniques of May [@May1972; @May1973; @Ahmadian2015]. In these analyses, row means of $\mathbf{A}$ are expected to be identical, but variation around this expectation will naturally arise due to random sampling of $\mathbf{A}$ off-diagonal elements and finite $S$. In simulations, the total variation in $\mathbf{M} = \mathbf{\gamma} \mathbf{A}$ row means that is attributable to $\mathbf{A}$ is small relative to that attributable to $\mathbf{\gamma}$, especially at high $S$. Variation in $\mathbf{\gamma}$ specifically isolates the effects of differing component response rates, hence causing differences in expected $\mathbf{M}$ row means.

### 2019-02-27 notebook/ms.tex
Edit the manuscript up to Reviewer 2 specific comment 3 -- these are next on the list, specifically the new Discussion paragraph

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} =\mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{23}}.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.724 and 0.732 for a and b,
respectively.

>>>>

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{23}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) row means that is
attributable to \(\mathbf{A}\) is small relative to that attributable to
\(\mathbf{\gamma}\), especially at high \(S\). Variation in
\(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.719 and 0.737 for a and b,
respectively.
